Chemical potential of liquids and mixtures via Adaptive Resolution Simulation

### 2014-05-20 writeDown/pub.jcp/chem-pot-jcp.tex
revise...

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}            
\newcommand{\bluec}[1]{{\color{blue} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{ex}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}
\newcommand{\typea}{{\textrm{A}}}
\newcommand{\typeb}{{\textrm{B}}}

@@

$w(x)$ (in green) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\vect R}_{c}$. The extension of this additional region is equal to the cutoff radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion \recheck{(see Eq.12 of Ref.\cite{prx})}}.

@@

Moreover we have shown that, because of the necessary conditions, the accuracy in the atomistic region is independent of the accuracy of the coarse-grained model, thus, in the coarse-grained region, one can use a generic liquid of spheres whose only requirement is that it has the same molecular density of the reference system.

In the simulation set up, ${\vect F}_{th}$ is calculated via an iterative procedure \recheck{using} the molecular number density in $\Delta$. The iterative scheme consists of calculating \recheck{${\vect F}^{k+1}_{th}({\vect r}) = {\vect F}^{k}_{th}({\vect r}) + \frac{1}{\rho^2_{0}\kappa_{T}}\nabla \rho^{k}({\vect r})$} ($\kappa_{T}$ the isothermal compressibility), and the thermodynamic force is considered converged when the target density $\rho_{0}$ is reached in $\Delta$. As a result, ${\vect F}_{th}({\vect r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force.
% proportional to the gradient of the corresponding higher order of the probability distribution;
\recheck{For example, the COM-COM radial distribution function correction for the second order~\cite{jctchan}.} Next it was proved that indeed the target Grand Canonical distribution, that is the probability distribution of a subsystem (of the size of the atomistic region in GC-AdResS) in a large full atomistic simulation is accurately reproduced. A large number of tests were performed and the reproduction by GC-AdResS of the probability distribution was numerically proved up to (at least) the third order, more than sufficient in MD simulations. Within this framework it was finally shown that the sum of work of ${\vect F}_{th}({\vect r})$ and that of the thermostat corresponds to the difference in chemical potential between the atomistic and coarse-grained resolution; this subject is treated in the next section.

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in \recheck{the original} AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the {Appendix~\ref{app:tmp2}})}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{\AT}$ and $U_{\CG}$ are the atomistic and coarse-grained potential.
\recheck{It must be noticed that the auxiliary Hamiltonian approach shall not be considered a Hamiltonian approach to adaptive resolution simulation. In fact, as discussed in Ref.\cite{prx} the equilibrium is imposed artificially and {\it per se} does not have any physical meaning (for more details, see discussion in the {Appendix~\ref{app:tmp2}})}. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

@@

\section{Analytic Derivation of the $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We consider a potential coupling between the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where  $\alpha$ and $\beta$ denotes the atom indexes of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$ of molecule $i$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

@@

\section{Results and Discussion}
We have calculated $\mu^{ex}$ for different liquids and mixtures, choosing cases which are representative of a large class of systems. Hydrophobic solvation in methane/water and in ethane/water mixtures, hydrophilic solvation in urea/water, a balance of both in water/tert-Butyl alcohol (TBA) mixture, other liquids, e.g. pure methanol and DMSO (and their mixtures with water), non aqueous mixtures in TBA/DMSO and alkane liquids such as methane, ethane and propane. Moreover, systems as water/urea are commonly used as cosolvent of biological molecules \cite{nico-debashish} while systems as tert-Butyl alcohol/water play a key role in modern technology \cite{irata}, thus they are of high interest {\it per se}. All technical details of each simulation are presented in the Appendix~\ref{app:tmp1}.
Results are reported in Table \ref{table} where the comparison with values obtained using full atomistic TI and available experiments, at the same concentrations, of our calculation is made; in our previous work we have already shown that value of the chemical potential of liquid water obtained with IPM is well reproduced by GC-AdResS, however the computational cost of IMP was very large, thus we do not consider calculations done with IPM in this paper. 
The agreement with full atomistic TI simulations is satisfactory in all cases, and thus it proves the solidity of GC-AdResS in describing the essential thermodynamics of a large class of systems.
We also compare the obtained values with those available in literature \cite{vang,nico}. Although the concentration of the minor component in the mixtures that we consider, is higher than the concentrations considered in Refs.\cite{vang,nico}, we are anyway in the very dilute regime and thus the chemical potential should not change in a significant way; we have verified  such a supposed consistency. 
The chemical potential of $k$-th liquid's component in a mixture is calculated as \recheck{(see Appendix~\ref{app:tmp3})}:
\begin{equation}
  \mu^{\exc,k}_{\AT}=
  \mu^{\exc,k}_{\CG}-
  \int_{\Delta}{\vect F}^k_{\thf}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r, k} d{\vect r}
% \mu^{ex,i}_{AT}=\mu^{ex,i}_{CG}-\int_{\Delta}{F}^{i}_{th}({\vect r})d{\vect r}-\int_{\Delta}\langle w\nabla_{x_{i}} w(x)(V_{AT}-V_{CG})\rangle_{\vect r}d{\vect r}
\label{eqn:tmp25}
\end{equation}
where ${F}^{k}_{\thf}(x)$ is the thermodynamic force applied to the molecules of the $k$-th component; this assures that, at the given concentration, the density of molecules of species $k$, in the transition region, is equivalent to the density of the same liquid's component in a reference full atomistic simulation.
\recheck{The ensemble average is taken over the position of the second molecule, provided that the first molecule is of species $k$, and located at position $\vect r$}.
% $\nabla_{x_{i}} w(x)$ is the gradient along $x$ (in a rectangular box) applied to the switching function $w(x)$ in the COM of molecules of component $i$.

\begin{table}[htpb]
\begin{center}
%\begin{tabular}{ccccc}
\begin{tabular}{ccccc}
\hline \hline
 Liquid component & Mole fraction of solute & GC-AdResS & TI & Experiment \\
\hline
water    & -- & -22.8 $\pm$ 0.2  & -22.1 $\pm$ 0.3 & -23.5 \cite{florian} \\
methane  & -- & -4.6 $\pm$ 0.1  & -5.2 $\pm$ 0.1 & -- \\
ethane   & -- & -8.2 $\pm$ 0.3  &  -8.8 $\pm$ 0.1 & -- \\
propane  & -- & -8.5 $\pm$ 0.1 & -9.5 $\pm$ 0.2 & -- \\
methanol  & -- & -20.1 $\pm$ 0.1 & -20.6 $\pm$ 0.4 & -20.5 \cite{vang}  \\
DMSO & -- & -32.2 $\pm$ 0.3 & -34.7 $\pm$ 0.7 & -32.2 \cite{dmso}  \\
methanol in methanol/water mixture & 0.01 & -18.1 $\pm$ 0.2 & -19.7 $\pm$ 0.2 & -- \\
methane in methane/water mixture & 0.006 & 9.1 $\pm$ 0.1  & 8.5 $\pm$ 0.2 & -- \\
urea in urea/water mixture & 0.02 & -56.1 $\pm$ 0.6 & -58.2 $\pm$ 0.5 & -57.8 $\pm$ 2.5 \cite{urea} \\
ethane in ethane/water mixture & 0.006 & 7.2 $\pm$ 0.2 & 7.4 $\pm$ 0.3 & -- \\ 
TBA in water/TBA mixture & 0.02 & -20.5 $\pm$ 0.6 & -21.4 $\pm$ 0.5 & -- \\
DMSO in DMSO/water mixture & 0.01 & -31.4 $\pm$ 0.5 & -33.2 $\pm$ 0.3 & -- \\
TBA in TBA/DMSO mixture & 0.02 & -24.8 $\pm$ 0.4 & -24.0 $\pm$ 0.5 & -- \\
\hline \hline
\end{tabular}
\caption{The excess chemical potential of different liquids and mixtures in $kJ/mol$ calculated from GC-AdResS and TI of full atomistic simulations. Experimental values for systems at the same concentrations used in simulation are also reported for comparison. For pure systems (water and methanol) we compare our values with those obtained in literature using the same force field and computational code. For mixtures, most of the values from literature (simulation and experiments) are available at lower concentrations (see Refs.\cite{vang,nico}); however since we are always in a very dilute regime the chemical potential does not change significantly; we have verified that such consistency holds. Note that the chemical potential of water in dilute mixtures is the same of pure water and is not reported above.
}
\label{table}
\end{center}
\end{table}

\begin{figure}
\center
\includegraphics[width=0.475\textwidth]{alcohol_rho.eps}\\
\includegraphics[width=0.475\textwidth]{water_rho.eps}
\caption{Top: Molecular density profile in $\Delta$ for TBA/water mixture; bottom, the same plot for water. Among all the systems considered, in this case the action of the thermodynamic force and that of the thermostat leads to the largest deviation from the reference all atomistic average density; however even in this case the discrepancy is negligible.\label{urea-TBA}}
\end{figure}
\begin{figure}
\center
\includegraphics[width=0.475\textwidth]{alcohol-alcohol.eps}\\
\includegraphics[width=0.475\textwidth]{alcohol-water.eps}\\
\includegraphics[width=0.475\textwidth]{water-water.eps}
\caption{Top: TBA-TBA radial distribution function; middle: the same plot for TBA-water; bottom: for water-water. The $g(r)$ is calculated only in the atomistic region.  Molecules located beyond a distance equal to the interactions' cut off, $0.9 nm$, are not considered since would include hybrid and coarse-grained molecules. \label{gr}}
\end{figure}

In essence, according to the results obtained, GC-AdResS allows an on-the-fly determination of $\mu^{ex}$ of each component of a liquid, whenever a simulation is performed, without extra computational costs.
Moreover, Fig.\ref{urea-TBA} shows the action of the thermodynamic force and of the thermostat in the transition region $\Delta$ for TBA-water; the molecular density is sufficiently close to that of reference (the largest difference is below $20\%$ and the average difference is below $10\%$), and thus it assures that in the atomistic region there are no (significant) artificial effects on the molecular density due to the perturbation represented by the interpolation of forces in $\Delta$. In Fig.\ref{gr} we report various radial distribution functions for TBA-water in the atomistic region of the adaptive set up. The agreement with data from a full atomistic simulation is highly satisfactory. Moreover, it must be underlined that, on purpose, we have chosen extreme technical conditions, that is, a very small atomistic and coarse-grained region ($0.5 nm$) and a relatively large transition region ($2.7 nm$). Even in these conditions we prove that local properties as those of Fig.\ref{urea-TBA} and Fig.\ref{gr}, together with a relevant thermodynamics quantity as $\mu^{ex}$ are well reproduced.
This example shows the key features of GC-AdResS, that is, a multiscale simulation where the chemical potential of each component is obtained without extra computational costs and with high accuracy in a simulation where other properties are also calculated with high accuracy. It must be also noticed that the system corresponding to the figures is, among all the system considered, the case where the action of the thermodynamic force and of the thermostat produces the less accurate agreement with the reference data.

>>>>

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}            
\newcommand{\bluec}[1]{{\color{blue} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{ex}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}
\newcommand{\typea}{{\textrm{A}}}
\newcommand{\typeb}{{\textrm{B}}}
\newcommand{\concenttba}{x_{\textrm{TBA}}}

@@

$w(x)$ (in green) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\vect R}_{c}$. The extension of this additional region is equal to the cut-off radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion \recheck{(see Eq.12 of Ref.\cite{prx})}}.

@@

Moreover we have shown that, because of the necessary conditions, the accuracy in the atomistic region is independent of the accuracy of the coarse-grained model, thus, in the coarse-grained region, one can use a generic liquid of spheres whose only requirement is that it has the same molecular density of the reference system. In the simulation set up, ${\vect F}_{th}$ is calculated via an iterative procedure \recheck{using} the molecular number density in $\Delta$. The iterative scheme consists of calculating \recheck{${\vect F}^{k+1}_{th}({\vect r}) = {\vect F}^{k}_{th}({\vect r}) + \frac{1}{\rho^2_{0}\kappa_{T}}\nabla \rho^{k}({\vect r})$} ($\kappa_{T}$ the isothermal compressibility), and the thermodynamic force is considered converged when the target density $\rho_{0}$ is reached in $\Delta$. As a result, ${\vect F}_{th}({\vect r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force.
% proportional to the gradient of the corresponding higher order of the probability distribution;
\recheck{For example, the COM-COM radial distribution function correction for the second order~\cite{jctchan}.} Next it was proved that indeed the target Grand Canonical distribution, that is the probability distribution of a subsystem (of the size of the atomistic region in GC-AdResS) in a large full atomistic simulation is accurately reproduced. A large number of tests were performed and the reproduction by GC-AdResS of the probability distribution was numerically proved up to (at least) the third order, more than sufficient in MD simulations. Within this framework it was finally shown that the sum of work of ${\vect F}_{th}({\vect r})$ and that of the thermostat corresponds to the difference in chemical potential between the atomistic and coarse-grained resolution; this subject is treated in the next section.

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in \recheck{the original} AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the {Appendix~\ref{app:tmp2}})}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{\AT}$ and $U_{\CG}$ are the atomistic and coarse-grained potential.
\recheck{It must be noticed that the auxiliary Hamiltonian approach shall not be considered a Hamiltonian approach to adaptive resolution simulation. In fact, as discussed in Ref.\cite{prx} the equilibrium is imposed artificially and {\it per se} does not have any physical meaning (for more details, see discussion in the {Appendix~\ref{app:tmp2}})}. 
\recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.\\

@@

\section{Analytic Derivation of  $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We consider a potential coupling between the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where  $\alpha$ and $\beta$ denotes the atom indexes of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$ of molecule $i$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

@@

\section{Results and Discussion}
We have calculated $\mu^{ex}$ for different liquids and mixtures, choosing cases which are representative of a large class of systems. Hydrophobic solvation in methane/water and in ethane/water mixtures, hydrophilic solvation in urea/water, a balance of both in water/tert-Butyl alcohol (TBA) mixture, other liquids, e.g. pure methanol and DMSO (and their mixtures with water), non aqueous mixtures in TBA/DMSO and alkane liquids such as methane, ethane and propane. Moreover, systems as water/urea are commonly used as cosolvent of biological molecules \cite{nico-debashish} while systems as tert-Butyl alcohol/water play a key role in modern technology \cite{irata}, thus they are of high interest {\it per se}. All technical details of each simulation are presented in the Appendix~\ref{app:tmp1}.\\

Results are reported in Table \ref{table}, where the comparison with values obtained using full atomistic TI and available experiments, at the same concentrations, of our calculation is made; in our previous work we have already shown that value of the chemical potential of liquid water obtained with IPM is well reproduced by GC-AdResS, however the computational cost of IMP was very large, thus we do not consider calculations done with IPM in this paper. 
The agreement with full atomistic TI simulations is satisfactory in all cases, and thus it proves the solidity of GC-AdResS in describing the essential thermodynamics of a large class of systems.
We also compare the obtained values with those available in literature \cite{vang,nico}. Although the concentration of the minor component in the mixtures that we consider, is higher than the concentrations considered in Refs.\cite{vang,nico}, we are anyway in the very dilute regime and thus the chemical potential should not change in a significant way; \recheck{we have verified  such a supposed consistency for one relevant system (see discussion about Fig.\ref{concentration})}. 
The chemical potential of $k$-th liquid's component in a mixture is calculated as \recheck{(see Appendix~\ref{app:tmp3})}:
\begin{equation}
  \mu^{\exc,k}_{\AT}=
  \mu^{\exc,k}_{\CG}-
  \int_{\Delta}{\vect F}^k_{\thf}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r, k} d{\vect r}
% \mu^{ex,i}_{AT}=\mu^{ex,i}_{CG}-\int_{\Delta}{F}^{i}_{th}({\vect r})d{\vect r}-\int_{\Delta}\langle w\nabla_{x_{i}} w(x)(V_{AT}-V_{CG})\rangle_{\vect r}d{\vect r}
\label{eqn:tmp25}
\end{equation}
where $\vect{F}^{k}_{\thf}(x)$ is the thermodynamic force applied to the molecules of the $k$-th component; this assures that, at the given concentration, the density of molecules of species $k$, in the transition region, is equivalent to the density of the same liquid's component in a reference full atomistic simulation.
\recheck{The ensemble average is taken over the position of the second molecule, provided that the first molecule is of species $k$, and located at position $\vect r$}.\\

\begin{table}[]
\begin{center}
%\begin{tabular}{ccccc}
\begin{tabular}{ccccc}
\hline \hline
 Liquid component & Mole fraction of solute & GC-AdResS & TI & Experiment \\
\hline
\recheck{water} & -- & -22.1 $\pm$ 0.2  & -- & -23.5 \cite{florian} \\
methane  & -- & -4.6 $\pm$ 0.1  & -5.2 $\pm$ 0.1 & -- \\
ethane   & -- & -8.2 $\pm$ 0.3  &  -8.8 $\pm$ 0.1 & -- \\
propane  & -- & -8.5 $\pm$ 0.1 & -9.5 $\pm$ 0.2 & -- \\
\recheck{methanol}  & -- & -20.3 $\pm$ 0.2 & -20.9 $\pm$ 0.3 & -20.5 \cite{vang}  \\
DMSO & -- & -32.2 $\pm$ 0.3 & -34.7 $\pm$ 0.7 & -32.2 \cite{dmso}  \\
\recheck{methanol in methanol/water mixture} & 0.01 & -19.0 $\pm$ 0.2 & -- & -- \\
\recheck{methane in methane/water mixture} & 0.006 & 9.3 $\pm$ 0.2  & -- & -- \\
urea in urea/water mixture & 0.02 & -56.1 $\pm$ 0.6 & -58.2 $\pm$ 0.5 & -57.8 $\pm$ 2.5 \cite{urea} \\
\recheck{ethane in ethane/water mixture} & 0.006 & 7.2 $\pm$ 0.2 & -- & -- \\ 
\recheck{TBA in water/TBA mixture} & 0.001 & -19.5 $\pm$ 0.5 & -20.8 $\pm$ 0.6 & -19.0 \cite{nico} \\
DMSO in DMSO/water mixture & 0.01 & -31.4 $\pm$ 0.5 & -33.2 $\pm$ 0.3 & -- \\
TBA in TBA/DMSO mixture & 0.02 & -24.8 $\pm$ 0.4 & -24.0 $\pm$ 0.5 & -- \\
\hline \hline
\end{tabular}
\caption{The excess chemical potential of different liquids and mixtures in kJ/mol calculated from GC-AdResS and TI of full atomistic simulations. Experimental values for systems at the same concentrations used in simulation are also reported for comparison. For pure systems (water and methanol) we compare our values with those obtained in literature using the same force field and computational code. For mixtures, most of the values from literature (simulation and experiments) are available at lower concentrations (see Refs.\cite{vang,nico}); However, since we are always in a very dilute regime the chemical potential does not change significantly. \recheck{We have provided evidence for the TBA/water misture that such consistency holds (see Fig.\ref{concentration})}. Note that the chemical potential of water in dilute mixtures is the same of pure water and is not reported above.
}
\label{table}
\end{center}
\end{table}

\begin{figure}
\center
\includegraphics[width=0.475\textwidth]{convergence.eps}
\caption{ \recheck{Excess chemical potential of Dimethyl sulfoxide (DMSO) in water as a function of cut-off radius calculated using GC-AdResS. The value obtained from thermodynamic integration calculation in also shown, with a gray region indicating the standard deviation. This value was calculated using a cut-off radius of $1.4$~nm. It was seen that the value does not change significantly if the cut-off radius is varied.}}
\label{convergence}
\end{figure}

\begin{figure}
\center
\includegraphics[width=0.475\textwidth]{concentration.eps}
\caption{ \recheck{Excess chemical potential of t-butyl alcohol (TBA) in water for different concentrations (in logarithmic scale), calculated using GC-AdResS. The results are compared with thermodynamic integration values. At mole-fraction $\concenttba=0.001$, the experiment value is shown. \label{concentration}}}
\end{figure}

\recheck{A complementary information to Table~\ref{table} is
  Fig.\ref{convergence}. In
  Table~\ref{table}, there is a systematic difference of
  $1.2 \sim 2.0$~kJ/mol between our values and those of TI
  calculations. In order to understand the possible origin of such a
  discrepancy and provide a way to eventually improve the accuracy of
  results, we have studied the behavior of $\mu^{ex}$ as a function of
  the interaction cut-off. In fact the current version of GC-AdResS,
  employs the reaction field method for treating electrostatic
  interactions in the atomistic region, and the cut off is likely to
  play a role. Fig.\ref{convergence}, for the case of DMSO/water
  mixture, confirms our intuition and suggests that we could
  systematically improve the accuracy by increasing the cut-off,
  however at a value of about $1.5$~nm, $\mu^{ex}$ converges. In any
  case, at a values of $1.4$~nm, which is the one routinely used in
  full atomistic simulations and used by us in all calculations, the
  value obtained with GC-AdResS is already satisfactory.\\

  A further
  question that may arise is the capability of our method to predict
  the behavior of $\mu^{ex}$ as a function of the concentration, above
  all in the very dilute regime. In Fig.\ref{concentration} we have
  performed such a study for the case of TBA/water mixture, we show a
  good agreement between GC-AdResS and TI and show that at a very
  dilute concentration our calculated value is close to that of
  experiments, moreover the trend, regarding the TI calculations, is
  consistent with that reported in Ref.\cite{nico}.\\}
